Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-06-12 WholePlotTrajectories.Rmd
discussion

keywords: [Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::gitbook: default
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::word_document2: default
---

@@

```{r Options, echo=FALSE, message=FALSE, warning=F}
knitr::opts_chunk$set(cache=F, warning=F, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40), fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")

@@

## A validation of the intermediate disturbance hypothesis

In accordance with the IDH the trajectories confirmed the increase of communities diversity after disturbance with the favored growth of otherwise less favored species.
However, although it was clear for the functional diversity this tendency was more blurry regarding communities taxonomic structure.

The taxonomic richness was enhanced after low disturbance but it was weakly or negatively impacted by intense disturbance, as observed on several post logging surveys [@Cannon1998; @Baraloto2012a].
The taxonomic evenness also increased after disturbance, but the correlation between the disturbance intensity and the diversity increase remained weak and valid only after 20 years (*i.e.* Simpson diversity).

Contrastingly the functional diversity increase after disturbance was determined by the disturbance intensity all along the 30 years monitored.
We already demonstrated that disturbance primarily entailed turnover of specific within communities, either among pre-disturbance survivors or among newly recruited trees.
It was demonstrated that the composition of old-growth survivors proved to mirror initial communities [@Herault2018], so disturbance rather impacted trees recruitment with the enhanced growth and survival of previously infrequent species and functional types.
Consistently disturbance resulted in an increase of the taxonomic dissimilarity between communities and their pre-disturbance composition, and in significant functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
Disturbance entailed a reorganization of the typical high dominance structure of hyperdiverse mature forests benefiting to pioneers and light demanding species: the changes in abiotic environment and competitive pressure favored pioneers which outcompete other species in non limiting resources but were excluded in mature forests by long-lived, more resistant and shade tolerant species.
Corroborating the IDH, the trajectories after disturbance relied on the environmental niches made available and filled by species from a restrictied functional range.
Recruited species therefore differed from pre-disturbance ones, constituting a community all the more diverse that the disturbance was intense [@Molino2001].

@@

Both the lag between taxonomic and functional recovery and the middling consistency of the IDH regarding communities taxonomic diversity were explained by the incomplete recovery of communities functional redundancy, which besides implies that communities resilience remained altered 30 years after disturbance [@Trenbath1999; @Elmqvist2003; @Diaz2005].

Functional redundancy mitigates the loss of species on communities functional structure [@Carmona2016], and therefore reduces the impact on disturbance.
The functional redundancy itself however might be reduced or re-organized within the functional space  and alter communities resilience.
Although the functional redundancy considered in the whole commnities functional space did not follow consistent trajectories after disturbance, the redundancy restricted to the functional space of the initial community clearly followed humped shaped trajectories.
This restricted redundancy first decreased, all the more so that the disturbance was intense, and then returned towards the initial value but this had not recovered for any disturbed community after 30 years.
After disturbance communities then differently occupied the functional space, as it was suggested by the shifts observed for the functional diversity and traits trajectories, before recovering the specific highly redundant initial functional space.
This increasing redundancy along the trajectory explained the hampered recovery of infrequent species specific to old-growth forests, as they were likely functionally redundant and therefore underwent competitive pressures limiting species functional similarity [@Mayfield2010].
The unachieved recovery of the functional redundancy meant a decreased resilience of pre-disturbance communities and an increased of disturbance-specific communities (corresponding to lower richness and higher dominance of pioneer species).
After disturbance the chances were therefore higher to have long lasting or self-maintained compositional changes in favor of disturbance resistant species, lianas or epiphytes [@Haddad2008; @Burslem2000; @Martin2013].
The recovery was not complete until the functional redundancy and resilience of the initial communities remained altered.
Specificaly, such incomplete recovery would threaten the maintenance of species contingent to undisturbed forests and increase the risk of cornerstone species loss and unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].

>>>>

keywords: [Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience]
journalinfo: Hal xxx
archive: DOI xxxx
date: "`r format(Sys.time(), '%d %B %Y')`"
preamble: >
  \usepackage{textcomp}
  \DeclareUnicodeCharacter{B0}{\textdegree}
  \usepackage{tabu}
  \usepackage{caption}
  \captionsetup{justification = justified}
  \renewenvironment{table}{\begin{table*}}{\end{table*}\ignorespacesafterend}
  \hyphenation{bio-di-ver-si-ty sap-lings}
bibliography: references.bib
lang: english
keywordlabel: Keywords
corrauthorlabel: Corresponding author
output:
  bookdown::pdf_book:
    base_format: EcoFoG::article
    keep_tex: TRUE
  bookdown::gitbook: default
  bookdown::word_document2: default
---

@@

```{r Options, echo=FALSE, message=FALSE, warning=F}
knitr::opts_chunk$set(cache=T, warning=F, tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=40), fig.align="center")
# Installation des packages si nécessaire et chargement
Library <- function(Packages) {
  InstallAndLoad <- function(Package) {
    if (!Package %in% installed.packages()[, 1]) {install.packages(Package, repos="https://cran.rstudio.com/")}
    require(Package, character.only = TRUE)
  }
  invisible(sapply(Packages, InstallAndLoad))
}
# Ajouter les packages nécessaires ici
Library("kableExtra")

@@

## The extent of the intermediate disturbance hypothesis

The examination of taxonomic and functional trajectories suggested that communities functional structure and dominant species pool relied upon deterministic processes following the IDH while the pool of rare species relied upon more stochastic processes.

The disturbance intensity did not predict the taxonomic richness, as already observed in the Paracou station [@Baraloto2012a] and in Bornean tropical forests [@Cannon1998], nor did it strongly correlate with the taxonomic evenness (*i.e.* Simpson diversity) although the it was enhanced after disturbance.

Contrastingly, the disturbance intensity predicted the increase in functional diversity along the 30 years monitored after disturbance.
The increase in functional diversity paralleled the important specific turnover and significant functional shifts towards resource-acquisitive strategies (sharp increase in the SLA, leaf thickness and bark thickness and decrease in wood density, leaf toughness and maximum height) [@Westoby1998; @Wright2004; @Reich2014].
As the old-growth, pre-disturbance survivor trees proved to mirror initial communities [@Herault2018] the functional trajectories and taxonomic turnover rather relied on the pool of newly recruited trees driven by the enhanced growth and survival of previously infrequent species and functional types.
Disturbance entailed a reorganization of the typical high dominance structure of hyperdiverse mature forests benefiting to pioneers and light demanding species: the changes in abiotic environment and competitive pressure favored pioneers which outcompete other species in non limiting resources but were excluded in mature forests by long-lived, more resistant and shade tolerant species.
Disturbance trajectories therefore relied on the environmental niches made available and filled by species from a restricted functional range that became dominant and structured the functionalcharacteristics of the community [@Grime1998].
The intermediate disturbance hypothesis predicted communities functional diversity and pool of dominant species [@Molino2001], but rare species relied on different processes and disprove the IDh for the whole taxonomic community [@Hubbell2001;@Chave2004].

@@

The functional and taxonomic decoupling and the disprove of the IDh for communities taxonomic structure were explained were explained by persisting changes in communities functoinal redundancy that implied communities incomplete resilience [@Trenbath1999;@Elmqvist2003;@Diaz2005].

Functional redundancy mitigates the loss of species on communities functional structure [@Carmona2016], and therefore reduces the impact on disturbance.
The functional redundancy itself however might be reduced or re-organized within the functional space  and alter communities resilience.
Although the functional redundancy considered in the whole communities functional space did not follow consistent trajectories after disturbance, the redundancy restricted to the functional space of the initial community clearly followed humped shaped trajectories.
This restricted redundancy first decreased, all the more so that the disturbance was intense, before returning towards the initial value but this had not recovered for any disturbed community after 30 years.
After disturbance communities differently occupied the functional space following traits trajectories and the shifts in functional composition, before recovering the initial functional space and its high redundancy.
This recovery phase was slow as it depend on infrequent, functionally redundant species specific to old-growth forests that rely upon different ecological processes, stochastic or deterministic competitive pressures limiting species functional similarity [@Chave2004;@Mayfield2010].

The unachieved recovery of the functional redundancy meant a decreased resilience of pre-disturbance communities and an increased of disturbance-specific communities (corresponding to lower richness and higher dominance of pioneer species).
After disturbance, chances were higher to observe long lasting or self-maintained compositional changes in favor of disturbance resistant species, lianas or epiphytes [@Haddad2008; @Burslem2000; @Martin2013].
The recovery remained incomplete until the recovery of the initial functional redundancy and resilience, and this specifically threatened the maintenance of species contingent to undisturbed forests and increased the risk of loosing cornerstone species and having unexpected ecological consequences [@Jones1994;@Chazdon2003a;@Diaz2005;@Gardner2007].

### 2018-06-12 WholePlotTrajectories.tex
discussion

\Abstract{
Understand the maintenance of forests biodiversity structure,
functioning and dynamics is urgent to anticipate their fate in the
global changing context. Communities diversity structure is assumed to
rely on a constant regime of disturbance and peak at intermediate
intensity. For tropical forests this intermediate disturbance hypothesis
(IDH) remain debated as well as the completeness of forests resilience
in all taxonomic and functoinal facets. To disentangle the ecological
processes driving forests dynamics we examined the the trajectories over
30 years after a gradient of logging and thinning disturbance, focusing
on communities taxonomic and functional composition and diversity.
Specifically we analysed the trajectories of communities taxonomic
richness and evennes and functional composition, diversity and
redundancy based on 7 key functional traits assessing species ecology.
We highlighted a decoupling between functional and taxonomic
trajectories, translating a fast recovery of communities functioning
while their taxonomic recovery was slowed by the recruitment of
infrequent species specific to mature forests. The IDH consistently
predicted communities functional trajectories but poorly reflected the
taxonomic ones, blurried by changes in the functional redundancy
entailing competitive exclusion. Although communities consistently
recovered their functioning after 30 years, initially infrequent species
proved long to recover which impacted the taxonomic structure and
functoinal redundancy over several decades. Our results acknowledged the
need of decades-long recovery cycles to restore the resilience of the
community and prevent species loss in the long term.
}

@@

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that give limited or misleading information on forests recovery
and functioning \citep{Martin2015, Chaudhary2016}. More releveant
monitoring would encompass communities composition, crucial for
conservation issues, and complete diversity profiles integrating both
species richness and evenness, to reveal the ecological rules
structuring communities \citep{Magurran1988, Lavorel2002, Bellwood2006}.
Furthermore, the functional approach that accounts for species
biological attributes and role in the ecosystem would be essential to
understand the correlations between ecosystems biodiversity, functioning
and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example in highlighting
deterministic processes in tropical rainforests that foster fast growing
species with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008, Ruger2009}. Communities then shifted from
a dominance of ``conservative'' slow-growing species dealing with scarce
resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift translated
into consistent trajectories for key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
and life history traits (seed mass and maximum height)
\citep{Wright2004, Westoby2006a, Chave2009b}.

A combination of taxonomic and functional approaches are essential to
fully assess communities response to disturbance and differences in
their respective trajectories are insightful of underlying mechanisms
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary and sometimes decouples, they are combined
in the measure of communities functional redundancy that quantifies the
amount of shared trait values among species \citep{Carmona2016}.
Functional redundancy is determinant of communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

\subsection{A validation of the intermediate disturbance
hypothesis}\label{a-validation-of-the-intermediate-disturbance-hypothesis}

In accordance with the IDH the trajectories confirmed the increase of
communities diversity after disturbance with the favored growth of
otherwise less favored species. However, although it was clear for the
functional diversity this tendency was more blurry regarding communities
taxonomic structure.

The taxonomic richness was enhanced after low disturbance but it was
weakly or negatively impacted by intense disturbance, as observed on
several post logging surveys \citep{Cannon1998, Baraloto2012a}. The
taxonomic evenness also increased after disturbance, but the correlation
between the disturbance intensity and the diversity increase remained
weak and valid only after 20 years (\emph{i.e.} Simpson diversity).

Contrastingly the functional diversity increase after disturbance was
determined by the disturbance intensity all along the 30 years
monitored. We already demonstrated that disturbance primarily entailed
turnover of specific within communities, either among pre-disturbance
survivors or among newly recruited trees. It wasdemonstrated that the
composition of old-growth survivors proved to mirror initial communities
\citep{Herault2018}, so disturbance rather impacted trees recruitment
with the enhanced growth and survival of previously infrequent species
and functional types. Consistently disturbance resulted in an increase
of the taxonomic dissimilarity between communities and their
pre-disturbance composition, and in significant functional shifts
towards resource-acquisitive strategies (sharp increase in the SLA, leaf
thickness and bark thickness and decrease in wood density, leaf
toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. Disturbance entailed a
reorganization of the typical high dominance structure of hyperdiverse
mature forests benefiting to pioneers and light demanding species: the
changes in abiotic environment and competitive pressure favored pioneers
which outcompete other species in non limiting resources but were
excluded in mature forests by long-lived, more resistant and shade
tolerant species. Corroborating the IDH, the trajectories after
disturbance relied on the environmental niches made available and filled
by species from a restrictied functional range. Recruited species
therefore differed from pre-disturbance ones, constituting a community
all the more diverse that the disturbance was intense
\citep{Molino2001}.

@@

Both the lag between taxonomic and functional recovery and the middling
consistency of the IDH regarding communities taxonomic diversity were
explained by the incomplete recovery of communities functional
redundancy, which besides implies that communities resilience remained
altered 30 years after disturbance
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Functional redundancy mitigates the loss of species on communities
functional structure \citep{Carmona2016}, and therefore reduces the
impact on disturbance. The functional redundancy itself however might be
reduced or re-organized within the functional space and alter
communities resilience. Although the functional redundancy considered in
the whole commnities functional space did not follow consistent
trajectories after disturbance, the redundancy restricted to the
functional space of the initial community clearly followed humped shaped
trajectories. This restricted redundancy first decreased, all the more
so that the disturbance was intense, and then returned towards the
initial value but this had not recovered for any disturbed community
after 30 years. After disturbance communities then differently occupied
the functional space, as it was suggested by the shifts observed for the
functional diversity and traits trajectories, before recovering the
specific highly redundant initial functional space. This increasing
redundancy along the trajectory explained the hampered recovery of
infrequent species specific to old-growth forests, as they were likely
functionally redundant and therefore underwent competitive pressures
limiting species functional similarity \citep{Mayfield2010}. The
unachieved recovery of the functional redundancy meant a decreased
resilience of pre-disturbance communities and an increased of
disturbance-specific communities (corresponding to lower richness and
higher dominance of pioneer species). After disturbance the chances were
therefore higher to have long lasting or self-maintained compositional
changes in favor of disturbance resistant species, lianas or epiphytes
\citep{Haddad2008, Burslem2000, Martin2013}. The recovery was not
complete until the functional redundancy and resilience of the initial
communities remained altered. Specificaly, such incomplete recovery
would threaten the maintenance of species contingent to undisturbed
forests and increase the risk of cornerstone species loss and unexpected
ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}.

>>>>

\Abstract{
Understand the maintenance of forests biodiversity structure,
functioning and dynamics is urgent to anticipate their fate in the
global changing context. Communities diversity structure is assumed to
rely on a constant regime of disturbance and peak at intermediate
intensity. For tropical forests this intermediate disturbance hypothesis
(IDH) remain debated as well as the completeness of forests resilience
in all taxonomic and functoinal facets. To disentangle the ecological
processes driving forests dynamics we examined the trajectories over 30
years after a gradient of logging and thinning disturbance, focusing on
communities taxonomic and functional composition and diversity.
Specifically we analysed the trajectories of communities taxonomic
richness and evennes and functional composition, diversity and
redundancy based on 7 key functional traits assessing species ecology.
We highlighted a decoupling between functional and taxonomic
trajectories, translating a fast recovery of communities functioning
while their taxonomic recovery was slowed by the recruitment of
infrequent species specific to mature forests. The IDH consistently
predicted communities functional trajectories but poorly reflected the
taxonomic ones, blurried by changes in the functional redundancy
entailing competitive exclusion. Although communities consistently
recovered their functioning after 30 years, initially infrequent species
proved long to recover which impacted the taxonomic structure and
functoinal redundancy over several decades. Our results acknowledged the
need of decades-long recovery cycles to restore the resilience of the
community and prevent species loss in the long term.
}

@@

In the short-term disturbance demonstrated negligible or even positive
impacts on communities diversity, which have been formalized by the
intermediate disturbance hypothesis (IDH) stating a maximized species
diversity at intermediate disturbance intensity
\citep{Molino2001, Kariuki2006a, Berry2008a}. Still, validations of the
IDH remain scarce in the long term and mainly rely on species richness
analyses that give limited or misleading information on forests recovery
and functioning \citep{Martin2015, Chaudhary2016}. More releveant
monitoring would encompass communities composition, crucial for
conservation issues, and complete diversity profiles integrating both
species richness and evenness, to reveal the ecological rules
structuring communities \citep{Magurran1988, Lavorel2002, Bellwood2006}.
Furthermore, the functional approach that accounts for species
biological attributes and role in the ecosystem would be essential to
understand the correlations between ecosystems biodiversity, functioning
and environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. The
functional trait-based approach focused on major traits related to
species ecology and performance was sucessfully adopted
\citep{Diaz2005, Villeger2008a}, for example in highlighting
deterministic processes in tropical rainforests that foster fast growing
species with efficient resources acquisition after disturbance
\citep{Molino2001, Haddad2008, Ruger2009}. Communities then shifted from
a dominance of ``conservative'' slow-growing species dealing with scarce
resources to ``acquisitive'' fast-growing species with rapid and
efficient use of abundant resources
\citep{TerSteege2001, Reich2014, Herault2011}. This shift translated
into consistent trajectories for key functional traits related to
resource acquisition (leaf area, density and chlorophyll content, and
stem specific gravity and bark thickness), tree growth and reproduction
life history traits (seed mass and maximum height)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.

A combination of taxonomic and functional approaches are essential to
fully assess communities response to disturbance and differences in
their respective trajectories are insightful of underlying mechanisms
\citep{Lohbeck2015, Guariguata2001}. Although taxonomic and functional
diversity are complementary and sometimes decoupled, they are combined
in the measure of communities functional redundancy that quantifies the
amount of shared trait values among species \citep{Carmona2016}.
Functional redundancy is determinant of communities resilience as high
redundancy, like in highly diverse tropical forests
\citep{Bellwood2006}, mitigates the impacts of species removal on
ecosystem functioning \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

@@

\subsection{The extent of the intermediate disturbance
hypothesis}\label{the-extent-of-the-intermediate-disturbance-hypothesis}

The examination of taxonomic and functional trajectories suggested that
communities functional structure and dominant species pool relied upon
deterministic processes following the IDH while the pool of rare species
relied upon more stochastic processes.

The disturbance intensity did not predict the taxonomic richness, as
already observed in the Paracou station \citep{Baraloto2012a} and in
Bornean tropical forests \citep{Cannon1998}, nor did it strongly
correlate with the taxonomic evenness (\emph{i.e.} Simpson diversity)
although the it was enhanced after disturbance.

Contrastingly, the disturbance intensity predicted the increase in
functional diversity along the 30 years monitored after disturbance. The
increase in functional diversity paralleled the important specific
turnover and significant functional shifts towards resource-acquisitive
strategies (sharp increase in the SLA, leaf thickness and bark thickness
and decrease in wood density, leaf toughness and maximum height)
\citep{Westoby1998, Wright2004, Reich2014}. As the old-growth,
pre-disturbance survivor trees proved to mirror initial communities
\citep{Herault2018} the functional trajectories and taxonomic turnover
rather relied on the pool of newly recruited trees driven by the
enhanced growth and survival of previously infrequent species and
functional types. Disturbance entailed a reorganization of the typical
high dominance structure of hyperdiverse mature forests benefiting to
pioneers and light demanding species: the changes in abiotic environment
and competitive pressure favored pioneers which outcompete other species
in non limiting resources but were excluded in mature forests by
long-lived, more resistant and shade tolerant species. Disturbance
trajectories therefore relied on the environmental niches made available
and filled by species from a restricted functional range that became
dominant and structured the functionalcharacteristics of the community
\citep{Grime1998}. The intermediate disturbance hypothesis predicted
communities functional diversity and pool of dominant species
\citep{Molino2001}, but rare species relied on different processes and
disprove the IDh for the whole taxonomic community
\citep{Hubbell2001, Chave2004}.

@@

The functional and taxonomic decoupling and the disprove of the IDh for
communities taxonomic structure were explained were explained by
persisting changes in communities functoinal redundancy that implied
communities incomplete resilience
\citep{Trenbath1999, Elmqvist2003, Diaz2005}.

Functional redundancy mitigates the loss of species on communities
functional structure \citep{Carmona2016}, and therefore reduces the
impact on disturbance. The functional redundancy itself however might be
reduced or re-organized within the functional space and alter
communities resilience. Although the functional redundancy considered in
the whole communities functional space did not follow consistent
trajectories after disturbance, the redundancy restricted to the
functional space of the initial community clearly followed humped shaped
trajectories. This restricted redundancy first decreased, all the more
so that the disturbance was intense, before returning towards the
initial value but this had not recovered for any disturbed community
after 30 years. After disturbance communities differently occupied the
functional space following traits trajectories and the shifts in
functional composition, before recovering the initial functional space
and its high redundancy. This recovery phase was slow as it depend on
infrequent, functionally redundant species specific to old-growth
forests that rely upon different ecological processes, stochastic or
deterministic competitive pressures limiting species functional
similarity \citep{Chave2004, Mayfield2010}.

The unachieved recovery of the functional redundancy meant a decreased
resilience of pre-disturbance communities and an increased of
disturbance-specific communities (corresponding to lower richness and
higher dominance of pioneer species). After disturbance, chances were
higher to observe long lasting or self-maintained compositional changes
in favor of disturbance resistant species, lianas or epiphytes
\citep{Haddad2008, Burslem2000, Martin2013}. The recovery remained
incomplete until the recovery of the initial functional redundancy and
resilience, and this specifically threatened the maintenance of species
contingent to undisturbed forests and increased the risk of loosing
cornerstone species and having unexpected ecological consequences
\citep{Jones1994, Chazdon2003a, Diaz2005, Gardner2007}.
